---
title: "In-text results"
output: html_document
date: "2023-09-08"
---
  
```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(english)
library(readxl)
library(baggr)

knitr::opts_chunk$set(include = FALSE)

here::i_am("code/generate_outputs/generate_text.Rmd")
source(here("code/functions/helpers.R"))
load(here("output/stan/bayesian-mr-models.Rdata"))
source(here("code/publication_bias/diarrhea-pub-bias.R"))
load(here("data/transformed/trials.Rdata"))
source(here("code/cea/cea-setup.R"))
default_gdp_pc <- read_rds(here("data/transformed/default_gdp_pc.rds"))
```

```{r headline bayesian results}
load(here("output/stan/bayesian-models-for-exhibits.Rdata"))
# overall bayesian estimates
bayes_or <- 
  oeb(bg_main) %>% 
  round(2) 

# bayesian estimate of reduction in mortality
bayes_reduction <- (1 - bayes_or["mean"]) * 100
```

```{r headline frequentist results}
source(here("code/ma_models/fit_ma_frequentist.R"))
# Overall frequentist estimates
freq_or <- 
  oef(fma_re, pval = FALSE) %>%
  round(2)

freq_reduction <- (1 - freq_or["mean"]) * 100
```

```{r headline cea}
df_cea <- read_rds(here("data/transformed/df_cea.rds"))
cost_per_daly <- df_cea %>% pull(cost_per_daly)
```


```{r number of studies}
df_studies <- read_rds(here("data/final/diarrhea_studies.rds"))

n_found <- df_studies %>% pull(reference) %>% n_distinct
n_included <- df_main_ma_adj %>% pull(trial) %>% n_distinct
n_waddington <- df_studies %>% dplyr::filter(source == "Waddington et al., 2023") %>% pull(reference) %>% n_distinct
n_waddington_included <- df_studies %>% dplyr::filter(source == "Waddington et al., 2023", included == 1) %>% pull(reference) %>% n_distinct
n_added <- df_studies %>% dplyr::filter(source == "New addition") %>% pull(reference) %>% n_distinct

n_mortality <- df_studies %>% dplyr::filter(mortality_reported == 1) %>% pull(reference) %>% n_distinct
```

```{r mortality reporting}
df_main_ma <-
  df_main_ma %>%
  mutate(mortality_reported = !str_detect(mortality_reporting, "Not reported"))

report_mortality <-
  df_main_ma %>% 
  dplyr::filter(mortality_reported) %>% 
  nrow

pre_registered <-
  df_main_ma_adj %>% 
  dplyr::filter(mortality_preregistered == 1) %>% 
  nrow

outcome <-
  df_main_ma_adj %>% 
  dplyr::filter(
    str_detect(mortality_reporting, "outcome")
  ) %>% 
  nrow

flow <-
  df_main_ma_adj %>% 
  dplyr::filter(
    str_detect(mortality_reporting, "Flow diagram")
  ) %>% 
  nrow

text <-
  df_main_ma_adj %>% 
  dplyr::filter(
    str_detect(mortality_reporting, "text")
  ) %>% 
  nrow

sample_reported <-
  df_main_ma_adj %>% 
  dplyr::filter(!str_detect(mortality_reporting, "Not reported")) %>% 
  mutate(N = c_n + t_n) %>% 
  pull(N) %>% 
  sum %>% 
  format(big.mark = ",")
```

```{r combined treatment}
always_combined <-
  df_main_ma_adj %>%
  dplyr::filter(treatment_arm_combined == 1) %>%
  nrow

some_combined <-
  df_main_ma_adj %>%
  dplyr::filter(treatment_arm_combined == 0, treatment_combined == 1) %>%
  nrow
```

```{r posterior prediction distribution}
bayes_ppd <- mint(exp(effect_draw(bg_main, 1e06))) %>% round(2)

next_implementation <- round(100*(1 - mean(bg_main_or_ppd)))

# Pr that effect is below 1 in next implementation:
round((sum(exp(effect_draw(bg_main, 1e06)) < 1)/1e06), 2)
```

# Abstract

We identified 52 RCTs and then obtained child mortality data from each study for which these data were collected and available, contacting authors of the study where necessary; this resulted in `r n_included` studies.
Frequentist and Bayesian methods were used to estimate the effect of water treatment on child mortality among included studies. We estimated a mean cross-study reduction in the odds of all-cause under-5 mortality of `r min(c(freq_reduction, bayes_reduction))`-`r max(c(freq_reduction, bayes_reduction))`% (frequentist odds ratio, OR, `r freq_or["mean"]`; 95% CI `r freq_or["lower"]` to `r freq_or["upper"]`; Bayes OR `r bayes_or["mean"]`; 95% CI `r bayes_or["lower"]` to `r bayes_or["upper"]`). The results were qualitatively similar under alternative modeling and data inclusion choices. Taking into account heterogeneity across studies, the expected reduction in a new study implementation is `r next_implementation`%.

After accounting for delivery costs, we estimate a cost per expected DALY averted due to water treatment between USD `r min(cost_per_daly)` and USD `r max(cost_per_daly)`, depending on approach.


# Results

## Systematic review

Figure 1 illustrates the search and the selection process. We identified 1,485 studies: 1,412 studies through databases and 73 studies included in (8) and (9). We screened these titles and abstracts to obtain a sample of `r n_found` studies for full-text review. **52** studies matched the inclusion criteria and we requested child mortality data from the authors of each study. **25** authors reported that they did not collect mortality data or that the data was no longer available. We screened additional `r n_waddington %>% english` studies, `r n_waddington_included %>% english` of which we included, based on another recent meta-analysis of safe drinking water interventions that was published after we performed our search (13). `r n_added %>% english` studies were added by us. 

The sample of `r ` studies with mortality data is summarized in Table 1. `r df_studies %>% dplyr::filter(mortality_reported == 1, included_dummy == 0) %>% pull(reference) %>% n_distinct %>% english %>% str_to_upper` studies were then excluded from the main analysis due to contamination in the control group but we conduct a sensitivity analysis with these studies included (see Supplementary Information, section 3).

## Publication bias

Neither Egger’s nor Andrews and Kasy’s tests provided evidence of publication bias on diarrhea or mortality outcomes. Based on the larger sample of `r n_found` studies, we also did not find evidence of the magnitude of effect or of positive significant effects on diarrhea being associated with availability of mortality outcomes. Since the power of these tests for mortality outcome may be limited when applied to our sample of `r n_included` studies, we also conducted post hoc simulations. We find that even if our search strategy has missed as many as **15** unpublished short studies with null effects (i.e. assuming mortality risk in both arms of **0.4%**, which is one quarter of annual mortality in our data) but, these were then found and added to our dataset, the meta-analytic estimate of OR would still be significant. 

## Characteristics of included studies

```{r total number of participants}
total_participants <-
  df_main_ma_adj %>% 
  mutate(N = c_n + t_n) %>% 
  pull(N) %>% 
  sum %>%
  format(big.mark = ",")
```

The studies included `r total_participants` participants. Out of `r n_included` studies, `r df_main_ma_adj %>% dplyr::filter(water_intervention == "Chlorination") %>% pull(trial) %>% n_distinct` were of water chlorination, `r df_main_ma_adj %>% dplyr::filter(water_intervention == "Filtration") %>% pull(trial) %>% n_distinct %>% english` of water filtration, and `r df_main_ma_adj %>% dplyr::filter(!(water_intervention %in% c("Filtration", "Chlorination"))) %>% group_by(water_intervention) %>% summarise(n = n_distinct(trial)) %>% pull(n) %>% unique %>% english` each of spring protection and solar disinfection. For `r df_main_ma_adj %>% dplyr::filter(data_source == "Microdata") %>% pull(trial) %>% n_distinct` studies we used individual-level data which were obtained from the authors or publicly available. In aggregate, `r tcases <- df_main_ma_adj %>% pull(tcases) %>% sum; tcases %>% format(big.mark = ",")` deaths occurred among `r t_n <- df_main_ma_adj %>% pull(t_n) %>% sum; t_n %>% format(big.mark = ",")` children in treatment arms (`r ((tcases/t_n) * 100) %>% round(1)`%); in the control arms, `r ccases <- df_main_ma_adj %>% pull(ccases) %>% sum; ccases %>% format(big.mark = ",")` deaths occurred among `r c_n <- df_main_ma_adj %>% pull(c_n) %>% sum; c_n %>% format(big.mark = ",")` children (`r ((ccases/c_n) * 100) %>% round(1)`%). `r df_main_ma_adj %>% dplyr::filter(ccases == 0 | tcases == 0) %>% pull(trial_name) %>% n_distinct %>% english` studies had no deaths in control and/or treatment arms. The annual risk of mortality in the pooled control group was about `r df_main_ma_adj %>% mutate(ccases_per_yr = 52*ccases/weeks) %>% summarise((sum(ccases_per_yr)/sum(c_n))*100) %>% round(1)`%.

Out of `r n_included` studies, `r report_mortality %>% english` reported mortality in publications. Of these, only `r outcome %>% english` clearly reported it as an outcome and `r flow %>% english` reported it in a participant flow diagram. Only `r pre_registered %>% english` study (23) clearly pre-registered mortality as an outcome.

We found studies to be representative of diarrhea prevalence in LMICs (see Figure S6) and found no significant differences to the larger set of RCTs which measured diarrhea. At the time of intervention, `r df_studies %>% dplyr::filter(included == "Included", income_group_hist == "Lower middle income") %>% nrow %>% english` countries were lower-middle income and the remaining were all low income. `r always_combined %>% english` studies included other interventions (`r df_main_ma_adj  %>% dplyr::filter(treatment_arm_combined == 1) %>% pull(treatment_arm) %>% str_remove_all("(Filtration) | (Chlorination) | (Floc) | (Bleach) | (Flocculant-disinfectant)") %>% str_remove_all("[+]") %>% str_squish %>% unique %>% paste(collapse = ",")`) and we conducted sensitivity analysis which excluded these studies. 


## Meta-analysis

```{r chlroination only estimates}
load(here("output/stan/bayesian-models-subsets.Rdata"))

# Chlorination only bayesian estimates
bayes_cl <- 
  oeb(bg_subsets$df_chlorination) %>% 
  round(2) 

# Chlorination only frequentist estimates
freq_cl <-
  oef(fma_subsets$df_chlorination, pval = FALSE) %>%
  round(2)
```

In the full set of `r n_included` studies, using a random-effects model we estimated a significant average reduction in odds of all-cause child mortality of `r freq_reduction`% (frequentist OR `r freq_or["mean"]`; CI 95% `r freq_or["lower"]`, `r freq_or["upper"]`) or `r bayes_reduction`% (Bayes OR `r bayes_or["mean"]`; CrI 95% `r bayes_or["lower"]`, `r bayes_or["upper"]`), depending on the model (see Figure 2).^[Since at low event rates ORs are approximately equal to RRs, assuming under-5 mortality in settings without access to clean water is **5% (see Table S7 for details), our Bayesian OR estimate implies mean risk reduction of 29% (Bayes RR of 0.71; 95% CrI 0.50, 0.92)**] Restricting the analysis to studies including chlorination, the reduction was `r (1 - freq_cl["mean"]) * 100`% and `r (1 - bayes_cl["mean"]) * 100`% respectively (frequentist OR `r freq_cl["mean"]`; CI 95% `r freq_cl["lower"]`, `r freq_cl["upper"]`; Bayes OR `r bayes_cl["mean"]`; CrI 95% `r bayes_cl["lower"]`, `r bayes_cl["upper"]`).


```{r}
fma_re %>% print(exp = TRUE)
```

```{r}
bg_main_full %>% print(exp=TRUE)
```

In Figure 2 we show a forest plot of study-level OR inputs into the analysis and the Bayesian and frequentist meta-analytic estimates. For posterior Bayesian estimates of effect in individual studies, mean ORs all fell within +-0.10 of the global mean. Despite strong pooling, there is mixed evidence on cross-study heterogeneity. The frequentist restricted maximum likelihood estimator suggested no heterogeneity *(tau = 0; SE = 0.05)*, which means its estimated effect is the same as in the fixed-effects model (as well as the Bayesian fixed effects model; *OR = 0.75, 95% CrI 0.60, 0.93*). However, frequentist meta-analyses can underestimate between-study variation and “snap” estimates to zero (14). Given the differences in the studied interventions and in the settings where the studies were carried out, a random-effects model is most appropriate. The Bayesian estimate of heterogeneity, in contrast, was considerable 

```{r}
bg_main
```

*(between-study SD of 0.24 on log scale, compared to the mean log(OR) effect of -0.33) but imprecisely determined (95% CrI 0.01, 0.73)*

Expected reduction in mortality odds in a new implementation, which is used by cost-effectiveness calculations and based on Bayesian posterior predictive distribution, was `r next_implementation`% (Bayes OR of `r bayes_ppd["mean"]`; 95% CrI `r bayes_ppd["2.5%"]`, `r bayes_ppd["97.5%"]`).

```{r sensitivity analyses}
all_sens_df <-
  bind_rows(
    freq_estimates_duration %>% 
      dplyr::select(mean, lower, upper),
    sub_estimates_bayes %>%
      dplyr::filter(
        !(rownames(sub_estimates_bayes) %in% c("df_chlorination", "df_filtration", "df_spring"))
      ) %>% 
      dplyr::select(mean, lower, upper),
    sub_estimates_freq %>%
      dplyr::filter(
        !(rownames(sub_estimates_freq) %in% c("df_chlorination", "df_filtration", "df_spring"))
      ) %>% 
      dplyr::select(mean, lower, upper),
    loo_freq %>% select(mean, lower, upper),
    loo_bayes %>% select(mean, lower, upper),
    oeb(bg_main_full),
    oef(fma_fe)
)
```

We provide results for all sensitivity analyses of data and model choices that we performed in Materials and Methods, section 3. The estimates from sensitivity analyses we attempted remained qualitatively similar, with mean OR estimates from `r nrow(all_sens_df)` models ranging from `r min(all_sens_df$mean) %>% round(2)` to `r max(all_sens_df$mean) %>% round(2)` for alternative data and modeling choices.

```{r diarrhea effects}
mr_diarrhea <- lapply(mr_fits, function(x) fixed_effects(x, s=T)[,,1]) %>% pluck("diarrhea_effects")
```

Univariate Bayesian meta-regression models found a significant relationship (in the sense of 95% interval not including zero) between effects on mortality and effects on diarrhea (beta = `r mr_diarrhea["mean"] %>% round(2)`, 95% CrI `r mr_diarrhea["lci"] %>% round(2)` to `r mr_diarrhea["uci"] %>% round(2)`), but not on compliance, prevalence of diarrhea, year of study, type of intervention, or randomisation unit (cluster vs individual).

## Cost-effectiveness

```{r cost-effectiveness}
df_cea_bigtab <-
  read_csv(
    here(
      "output/tables/table-cea-estimates.csv"
    )
  ) %>%
  tibble::column_to_rownames(var = "...1")
```

The cost-effectiveness calculations are based on the expected `r next_implementation`% reduction in the odds of mortality in a new implementation.

### Point-of-access

We calculate the per-child cost of provision of chlorine treated drinking water for five years at USD `r df_cea_bigtab["cost", "DSW"]`. This leads to cost per DALY averted by the dispenser program of USD `r df_cea_bigtab["cost_per_daly", "DSW"]` (Table 2, Column 1), far lower than Kenya’s GDP per capita (USD 2,099 in 2022). We estimate that chlorine dispensers program saves `r (10000/ df_cea_bigtab["cost_per_daly", "DSW"]) %>% plyr::round_any(10)` DALYs per $10,000 spent.


### Inline

We use cost estimates from India, where state and national governments have expanded access to piped water in rural areas, and at least two state governments are planning to incorporate inline chlorination, with technical assistance from the NGO Evidence Action. We assume an increase in chlorination of `r df_cea_bigtab["effective_tkup", "ILC"]*100`% (observed in a study of inline chlorination in Bangladesh.) The estimated per-child cost for five years is about USD `r df_cea_bigtab["cost", "ILC"] %>% plyr::round_any(10)` (Table 2 column 2). Based on this, we calculate `r df_cea_bigtab["cost_per_daly", "ILC"]` per DALY averted (Table 2 column 2). Again this is substantially lower than the 1x GDP threshold for India (USD 2,388 in 2022). We estimate that this program saves around `r (10000/ df_cea_bigtab["cost_per_daly", "ILC"]) %>% plyr::round_any(10)` DALYs per $10,000 spent.

### Coupons

A 150-milliliter bottle of dilute chlorine solution sufficient for treating one household-month of water costs USD 0.31 and we estimate the total program costs per-child at USD `r df_cea_bigtab["cost", "coupons"] %>% round` for five years (Table 2, column 3). This implies a cost of USD `r df_cea_bigtab["cost_per_daly", "coupons"]`  per DALY averted. This is substantially lower than the weighted average of GDP per capita for these countries (USD 2,200). MCH delivery appears to have lower cost per DALY than the other examples discussed here, in part because our analysis only considers the benefits of reduced child mortality and, unlike inline chlorination or dispensers, the coupon program specifically targets children under-5. We estimate that this program saves around `r (10000/ df_cea_bigtab["cost_per_daly", "coupons"]) %>% plyr::round_any(10)` DALYs per $10,000 spent. 

## Discussion

Randomized control trials studying the impact of water treatment are typically not individually powered for mortality. Aggregating data from `n_included` studies, we estimate that water treatment reduced the odds of all-cause child mortality by about a quarter on average. 


### Cost-effectiveness thresholds and lists of priority health interventions

The above cost-effectiveness estimates of \$`r min(cost_per_daly)`-\$`r max(cost_per_daly)` per DALY averted would place water treatment near the top of these lists. 

```{r GDP thresholds}
gdp_threshold_times <-
  round(rev(default_gdp_pc)/df_cea$cost_per_daly)

benken <- Inf
rr_check <- .999
while(benken > default_gdp_pc[["dsw"]]){
  benken <- cea_v2(result = "table", 
                   input="rr", rr_check,
                   p1=.0692, 
                   tkup=0.51, 
                   cost=dsw_cost, 
                   u5_per_hh=1,
                   gdp_pc=default_gdp_pc[["dsw"]])$cost_per_daly
  rr_check <- rr_check - .001
}

ce_threshold <- (100*(1-rr_check)) %>% round(2)
```

Our estimates suggest that water treatment exceeds the 1x GDP threshold `r min(gdp_threshold_times)`-`r max(gdp_threshold_times)` times. Indeed, because the cost of water treatment is low, even small effects on mortality would meet these thresholds. 

For example, repeating the calculation for chlorine dispensers in Kenya, we find that the threshold of 1x GDP is reached at `r ce_threshold` reduction in odds of under-5 mortality.

### Magnitude of benefits 

```{r global benefits}
s6 <- read_csv(
  here("output/tables/cea-globalbenefits.csv")
)

s6_total <- s6 %>% dplyr::filter(region == "Total")
```


To illustrate the potential magnitude of the benefits of water treatment, in Table S6 we present a back-of-envelope calculation, which suggests that a global coupon program delivering water treatment to all households with children under-5 who do not yet have access to piped water in low- and lower-middle income countries, could save over around `r (plyr::round_any(s6_total$u5_saved, 50) * 1000) %>% format(scientific  = FALSE, big.mark = ",")` under-five lives at a cost of approximately USD `r plyr::round_any(s6_total$coupons_cost, 50)` million each year

### Comparison with other sources of evidence

To illustrate, we calculate a simple model in which diarrheal deaths are taken from the central estimate of the Global Burden of Disease (GBD) project (2), the effect of water treatment on diarrhea is taken from the central estimate in an earlier meta-analysis (8), and mortality is assumed to be linear in diarrhea cases, so that reductions in diarrhea deaths are proportional to reductions in diarrheal cases. The point estimate from this model is a **3.9**% reduction in mortality risk. 
 
Suppose that an analyst puts some weight on the meta-analysis result and the rest of weight on the alternative model. We find that to meet the most-demanding cost-effectiveness threshold that we identified (USD 200 per DALY averted) the analyst needs to put only **6**% weight on the meta-analysis result in the case of chlorine dispensers (reaching the $200/DALY threshold at OR = **0.947**) and **16**% for inline chlorination (OR = **0.924**).
 
A formal way of combining these estimates in a way that takes model uncertainty into account is to use the alternative model as a Bayesian prior (centered at **3.9**% reduction) for our meta-analysis. We find that the water treatment meets USD 200 per DALY averted threshold unless the analyst’s prior is over **16** times more precise than the result from the meta-analysis alone for the case of dispensers and over **7** times more precise for inline chlorination. (For coupons any precision will suffice, since **3.9**% reduction meets the threshold.)
 
 ### Limitations

```{r cea and takeup}
cea_tkup <- function(...) {
  cea_v2(x = mean(default_or_ppd), input = "or", ...)$cost_per_daly
}

cea_compliance_df <- rbind(
  data.frame(method = "dsw",
             p1=.0692, 
             cost=dsw_cost, 
             u5_per_hh=1,
             gdp_pc=default_gdp_pc[["dsw"]]),
  data.frame(method = "ilc",
             p1 = 0.039,
             cost = 59.57/5,
             u5_per_hh=1,
             gdp_pc=default_gdp_pc[["ilc"]]),
  data.frame(method = "coupons",
             p1 = u5_mr$avg_mortality_rate,
             cost = cost_vouchers,
             u5_per_hh=u5_per_hh_coupons[["x"]],
             gdp_pc=default_gdp_pc[["coupons"]])
) %>% 
  expand_grid(tkup = seq(.001, .5, .0005)) %>%
  mutate(result = "table") %>% 
  mutate(res = pmap_dbl(select(., -method), cea_tkup)) %>% 
  dplyr::filter(gdp_pc >= res) %>% 
  group_by(method) %>% 
  summarise(min(tkup)) 
```

However, the effective take-up in water treatment needed to reach 1x GDP per capita threshold ranges from `r cea_compliance_df %>% dplyr::filter(method == "coupons") %>% pull("min(tkup)") %>% round(3) * 100`% (for coupons) to `r cea_compliance_df %>% dplyr::filter(method == "ilc") %>% pull("min(tkup)") %>% round(3) * 100`% (for ILC). We discuss the drivers of cost effectiveness further in Supplementary Information, section 4. 

### Limitations

However, we incorporate heterogeneity into our cost-effectiveness assessment by using predicted mean effect, which has a higher OR (smaller effect) than the mean within the `r n_included` studies.

For example, our main estimate for dispensers uses compliance data from NGO Evidence Action. However, data from other implementations suggests lower effects of dispenser provision on water treatment, as take-up might decline over time (32), or counterfactual take-up might be higher than expected (36). We also conduct cost-effectiveness analyses using alternative, lower compliance values for each delivery method (Table S7). These alternative estimates of cost per DALY are inversely proportional to drop in compliance: \$`r df_cea_bigtab["cost_per_daly", "DSW_low"]` for dispensers, \$`r df_cea_bigtab["cost_per_daly", "ILC_low"]` for ILC, and \$`r df_cea_bigtab["cost_per_daly", "coupons_low"]` for MCH delivery. These are not directly comparable since some are based on real compliance data and others hypothetical.

However, in terms of diarrhea prevalence we find that the studies included in this meta-analysis are broadly representative of the settings in which policymakers might implement water treatment programs in terms of diarrhea prevalence and there are no significant differences from a larger sample of **73** studies (Supplementary Information, section 1). 

### Lessons

Out of the `r report_mortality` studies that reported mortality in publications, `r outcome` report it as an outcome, `r flow` include number of deaths in a flow diagram, and `r text` discuss or report death in the text. Only `r pre_registered` includes mortality as an outcome in the pre-registration. Relying on reported data only would reduce the total sample size from `r total_participants` to `r sample_reported`. 

# Supplement

## 1 Details of included studies and comparison with other RCTs

### Summary of 18 included studies

```{r diarrhea ecdf}
diarrhea_ecdf <-
  read_csv(here("data/raw/IHME_GLOBAL_DIARRHEA_2000_2019_PREV_A1_S3_ADMIN_1_Y2020M08D31.CSV")) %>%
  select(ADM1_NAME, mean) %>%
  unique %>%
  # multiply by 5, since the data are entered as a rate
  # i.e. what is the prevalence of diarrhea among children of a given age, among u5s
  # rather than the prevalence of diarrhea among all children under 5
  mutate(
    mean = 5 * mean
  ) %>%
  pull(mean) %>%
  ecdf
```

The median follow-up length for mortality was `r median(df_main_ma_adj$weeks)` weeks, with the longest follow-up being 4-6 years, in two studies conducted in Kenya  (1). `r df_studies %>% dplyr::filter(included == "Included", income_group_hist == "Lower middle income") %>% pull(trial_name) %>% n_distinct %>% english`  studies were conducted in lower-middle income countries, and `r df_studies %>% dplyr::filter(included == "Included", income_group_hist == "Low income") %>% pull(trial_name) %>% n_distinct %>% english`  were conducted in low-income countries, according to the World Bank classification at the time of the study (2). Since then, `r df_studies %>% dplyr::filter(included == "Included", income_group_hist == "Lower middle income", income_group == "Upper middle income") %>% pull(country) %>% n_distinct %>% english` lower-middle income country (`r df_studies %>% dplyr::filter(included == "Included", income_group_hist == "Lower middle income", income_group == "Upper middle income") %>% pull(country) %>% unique %>% paste(collapse = ", ")`) where `r df_studies %>% dplyr::filter(included == "Included", income_group_hist == "Lower middle income", income_group == "Upper middle income") %>% pull(trial_name) %>% n_distinct %>% english` studies were conducted has become an upper-middle income country, and `r df_studies %>% dplyr::filter(included == "Included", income_group_hist == "Low income", income_group == "Lower middle income") %>% pull(country) %>% n_distinct %>% english` low income country (`r df_studies %>% dplyr::filter(included == "Included", income_group_hist == "Low income", income_group == "Lower middle income") %>% pull(country) %>% unique %>% paste(collapse = ", ")`) where `r df_studies %>% dplyr::filter(included == "Included", income_group_hist == "Low income", income_group == "Lower middle income") %>% pull(trial_name) %>% n_distinct %>% english` studies were conducted have become lower-middle income.

Out of the `r n_included` studies, `r n_rural <- df_studies %>% dplyr::filter(setting == "rural", included == "Included") %>% pull(reference) %>% n_distinct; n_rural` were conducted in rural areas, `r df_studies %>% dplyr::filter(setting == "mixed", included == "Included") %>% pull(reference) %>% n_distinct %>% english` were conducted in both rural and urban areas, and `r df_studies %>% dplyr::filter(setting == "urban", included == "Included") %>% pull(reference) %>% n_distinct %>% english` were conducted in a peri-urban setting. The compliance rate (see Table 1 for definition) in the sample ranged from a low of `r df_studies %>% pull(compliance) %>% min(na.rm = TRUE) * 100`% to a high of `r df_studies %>% pull(compliance) %>% max(na.rm = TRUE) * 100`%, with a median of `r round(median(df_studies$compliance, na.rm = TRUE) * 100, 0)`%.  Average use of water cleaning treatment (see later for definitions) in treatment groups was `r round(mean(df_studies$compliance, na.rm = TRUE) * 100, 0)`% and in control groups it was about `r df_main_ma_adj %>% select(trial_name, takeup_control) %>% mutate(takeup_control = takeup_control * 100) %>% unique %>% pull(takeup_control) %>% mean(na.rm = TRUE) %>% round(0)`%. Based on this, we assumed an effective compliance across meta-analysis (used for cost-effectiveness calculations) of **53%**.


Contamination level measures are not consistently reported across studies. **Four studies report 54 to 181 TTC/100 ml (thermotolerant coliforms, which include E. Coli and three other bacteria species). Another 4 studies report E.Coli concentration from 34 to 98 per 100 ml.**

Estimates of diarrhea prevalence among the `r n_included` included studies are representative of prevalence in low- and middle-income countries. Household surveys across 94 low- and middle-income countries found diarrhea prevalence in 2017 ranging from 0.2% to 56.7% across sub national units, with a median of 13.7% (3). Diarrhea prevalence rates (at baseline or, if baseline not available, in the control group) in our sample of studies range from `r round(min(df_main_ma$prevalence, na.rm = TRUE) * 100, 0)`% to `r round(max(df_main_ma$prevalence, na.rm = TRUE) * 100, 0)`%, with a weighted mean (using meta-analytic weights from the frequentist model) of `r prev <- summarise_prevalence_compliance$prevalence1; round(prev * 100, 1)`%; this corresponds to the `r round(diarrhea_ecdf(prev) * 100, 0)`st percentile of the distribution of sub national diarrhea estimates, see Figure S6. 

```{r diarrhea prevalence}
df_diarrhea_prev <-
  read_csv(here("data/raw/IHME_GLOBAL_DIARRHEA_2000_2019_PREV_A1_S3_ADMIN_1_Y2020M08D31.CSV"))

df_all_ma_adj_prev <- 
  df_all_ma_adj %>%
  select(trial_name, prevalence) %>%
  distinct() %>%
  rename(
    obs = trial_name, 
    mean = prevalence
  ) %>%
  mutate(group = "Studies")

df_diarrhea_prev <- 
  df_diarrhea_prev %>%
  select(ADM1_NAME, mean) %>%
  distinct() %>%
  rename(obs = ADM1_NAME) %>%
  # multiply by 5, since the data are entered as a rate
  # i.e. what is the prevalence of diarrhea among children of a given age, among u5s
  # rather than the prevalence of diarrhea among all children under 5
  mutate(
    mean = 5 * mean,
    group = "IHME"
  )

df_prev <- rbind(df_all_ma_adj_prev, df_diarrhea_prev)

x <- subset(df_prev, group == "IHME")$mean
x <- x[!is.na(x)]
percentile <- sum(x < summarise_prevalence_compliance$prevalence1)/length(x)
```

Diarrhea prevalence rates (at baseline or, if baseline not available, in the control group) in our sample of studies range from `r df_main_ma_adj %>% summarise(min(prevalence)) * 100`% to`r df_main_ma_adj %>% summarise(max(prevalence)) * 100`%, with a weighted mean (using weights from the frequentist model) of `r (summarise_prevalence_compliance["prevalence1"]* 100) %>% round(1)`%; this corresponds to the `r round(percentile * 100, 0)` percentile of the distribution of sub national diarrhea estimates (see Figure 2). 

### Choice of treatment and control groups 

For **fifteen** out of **eighteen* 18** studies, water treatment was compared to a pure control group which received no intervention. In three two of these, several experimental arms with different kinds of water treatment were combined. These included some combination of water chlorination, flocculant-disinfection, and safe storage vessels (Reller et al. 2003, Luby et al. 2006, Crump et al. 20054, 5). 

In `r always_combined %>% english` out of the `r n_included %>% english` cases, water treatment was always combined with another intervention: cookstoves in Kirby et al 2019 (6), other sanitation and hygiene interventions in Humphrey et al 2019 (7), andor safe storage in Chiller et al 2006, Peletez et al. 2012 (8), and Semenza et al 2018 (9). These are removed in sensitivity analyses, see Section 3. Quick et al. 1999 also combined safe storage and water treatment. However, the control group also received safe storage containers. With the exception of one treatment arm in Dupas et al 2023, all other interventions involved some sort of educational or promotion intervention instructing study participants on how to treat water with the relevant method, so education is not considered a combined intervention.

`r some_combined %>% english` studies have both a pure water treatment arm and combined intervention arms: Luby et al. 2006 includes an arm combining water treatment and handwashing; Luby et al. 2018 and Null et al. 2018 include a water, sanitation, and handwashing arm as well as a water, sanitation, handwashing, and nutrition arm; Reller et al 2003 combined water treatment and safe storage.


### Comparison of characteristics between included and excluded studies 

```{r inclusion exclusion analysis}
source("generate_incexcl_analysis.R")

pvals <-
  lapply(reg, function(x) summary(x)$coef[2, "Pr(>|t|)"]) %>% 
  unlist() %>% round(2)

n_unimproved <- 
  df_studies %>% 
  dplyr::filter(
    unimproved,
    included == "Included"
  ) %>% 
  pull(reference) %>% 
  n_distinct

n_rural <-
  df_studies %>%
  dplyr::filter(included == "Included", rural) %>%
  pull(reference) %>%
  n_distinct

n_nonrural <-
  df_studies %>%
  dplyr::filter(included == "Included", !rural) %>%
  pull(reference) %>%
  n_distinct


studies_wolf <- 
  df_studies %>%
  dplyr::filter(in_wolf_et_al) %>%
  pull(reference) %>%
  n_distinct()

n_wolf <- 
  df_studies %>%
  dplyr::filter(in_wolf_et_al) %>%
  nrow

studies_wolf_include <- 
  df_studies %>%
  dplyr::filter(in_wolf_et_al, included == "Included") %>%
  pull(reference) %>%
  n_distinct()

n_wolf_include <- 
  df_studies %>%
  dplyr::filter(in_wolf_et_al, included == "Included") %>%
  nrow

studies_wolf_rural <-
  df_studies %>%
  dplyr::filter(in_wolf_et_al, setting == "rural") %>%
  pull(reference) %>%
  n_distinct()

studies_wolf_mixed <-
  df_studies %>%
  dplyr::filter(in_wolf_et_al, setting == "mixed") %>%
  pull(reference) %>%
  n_distinct()

studies_wolf_urban <-
  df_studies %>%
  dplyr::filter(in_wolf_et_al, setting == "urban") %>%
  pull(reference) %>%
  n_distinct()

studies_unimproved_wolf <-
  df_studies %>%
  dplyr::filter(in_wolf_et_al, unimproved) %>%
  pull(reference) %>%
  n_distinct
```

We additionally compare some key characteristics of the water treatment studies included with those excluded from the analysis, but included in (8). There were `r studies_wolf` studies in (8), yielding `r n_wolf` observations. Some studies had multiple observations on account of multiple study locations, and hence yielded multiple effect estimates. `r studies_wolf_include` of these studies were included in our meta-analysis, resulting in `r n_wolf - n_wolf_include` observations excluded from our meta-analysis.

The distribution of effect estimates of water treatment on diarrhea and compliance rates are similar across included and excluded data (see Fig. S2).
`r studies_wolf_rural` out of `r studies_wolf` studies (`r round((studies_wolf_rural/studies_wolf)* 100, 1)`%) were conducted in rural settings, with `r round((studies_wolf_mixed/studies_wolf)* 100, 1)`% and `r round((studies_wolf_urban/studies_wolf)* 100, 1)`% being conducted in mixed and urban settings respectively. Similar to this, `r n_rural` out of the `r n_included` included studies (`r round((n_rural/n_included)*100)`%) are set in rural areas and `r n_nonrural` out of the `r n_included` (`r round((n_nonrural/n_included)*100)`%) included studies are conducted in mixed and urban settings.

In terms of the water source, the primary source of water at baseline (or in the control group) was an unimproved water source in `r studies_unimproved_wolf` out of `r studies_wolf` observations (`r round((studies_unimproved_wolf/studies_wolf)*100, 0)`%.) This is comparable to `r round((n_unimproved/n_included)*100,1)`% (`r n_unimproved` out of `r n_included` studies) among the included studies.  
A t-test of mean difference between included and excluded studies yield insignificant differences for the diarrhea effect size (p-value = `r pvals %>% pluck("effect on diarrhea")`), compliance rate (p-value = `r pvals %>% pluck("compliance")`), setting (binary variable indicating whether the setting was rural, p-value = `r pvals %>% pluck("rural")`), and presence of improved water sources at baseline, as classified in (8) (p-value = `r pvals %>% pluck("unimproved")`). We note that this final p-value is low. However, since we present the results from multiple tests, we consider several multiple testing corrections. Under the Bonferroni correction the revised p-value cutoffs for each test would be 0.025 at 10% significance. (Alternatively, the stepwise method due to Holm provides revised cutoffs of 0.033, 0.05, 0.1 and 0.025, respectively.)



### Age characteristics of included children

```{r}
df_main_ma %>%
  select(
    trial_name,
    obs, 
    weeks
  )
```


## 2 Meta-analysis models

### Comparing odds ratios and risk ratios for rare events

## 3 Sensitivity analysis


### Range of values in sensitivity analyses

```{r values in SA}
all_sens_df <-
  bind_rows(
    freq_estimates_duration %>% 
      dplyr::select(mean, lower, upper),
    sub_estimates_bayes %>%
      dplyr::filter(
        !(rownames(sub_estimates_bayes) %in% c("df_chlorination", "df_filtration", "df_spring"))
      ) %>% 
      dplyr::select(mean, lower, upper),
    sub_estimates_freq %>%
      dplyr::filter(
        !(rownames(sub_estimates_freq) %in% c("df_chlorination", "df_filtration", "df_spring"))
      ) %>% 
      dplyr::select(mean, lower, upper),
    loo_freq %>% select(mean, lower, upper),
    loo_bayes %>% select(mean, lower, upper),
    oeb(bg_main_full),
    oef(fma_fe)
)
```

Over a set of all sensitivity analyses (`r n_models <- nrow(all_sens_df); n_models` models) we found that  the study estimates remain qualitatively similar to our main estimate. In this set of sensitivity analyses, the mean OR estimates range from `r min(all_sens_df$mean) %>% round(2)` to `r max(all_sens_df$mean) %>% round(2)`. `r sum(all_sens_df$upper < 1)` out of `r n_models` models have upper 95% intervals below one.


### Case-by-case details of sensitivity analyses


```{r SA details}
s4 <- 
  read_csv(
    here(
      "output/tables/table-loo-study.csv"
    )
  ) %>%
  separate(
    CI_bayes,
    sep = ", ",
    into = c("CI_bayes_lb", "CI_bayes_ub")
  ) %>%
  separate(
    CI_freq,
    sep = ", ",
    into = c("CI_freq_lb", "CI_freq_ub")
  ) %>%
  mutate(
    across(
      matches("CI_(.*)_lb"),
      ~ . %>%
        str_sub(2, length(.)) %>%
        as.numeric
    ),
    across(
      matches("CI_(.*)_ub"),
      ~ . %>%
        str_sub(end = -2) %>%
        as.numeric
    )
  ) 
```

The frequentist and Bayesian odds ratio estimates are given in Table S4.  For the Bayesian model, the mean OR ranged from `r min(s4$mean_bayes) %>% round(2)` to `r max(s4$mean_bayes) %>% round(2)`, with the lowest 95% CrI  lower bound of `r min(s4$CI_bayes_lb)` and the highest 95% CrI upper bound of `r max(s4$CI_bayes_ub)`. For frequentist OR, the means ranged from `r min(s4$mean_freq) %>% round(2)` to `r max(s4$mean_freq) %>% round(2)`, with the lowest lower bound of `r min(s4$CI_freq_lb)` and the highest upper bound of `r max(s4$CI_freq_ub)`.


```{r more SA details}
s5 <- 
  read_csv(
    here(
      "output/tables/additional-sa-results.csv"
    )
  ) 

s5_means <- 
  s5 %>%
  slice(1) %>%
  mutate(
    across(
      everything(),
      ~as.numeric(.)
    )
  )

mean_freq <- s5_means %>% 
  dplyr::select(contains("Freq")) %>%
  rowMeans()

mean_bayesian <- s5_means %>% 
  dplyr::select(contains("Bayesian")) %>%
  rowMeans()
```

The meta-analysis estimates remain quantitatively similar and significant with a mean reduction in mortality odds of `r round((1-s5_means[1,2])*100)`-`r round((1-s5_means[1,1])*100)`%, depending on the model (see Table S5). 


Adding the solar disinfection trial (14) to the meta-analysis results in a mean reduction in mortality odds of `r round((1-s5_means[1,3])*100)`-`r round((1-s5_means[1,4])*100)`%; adding the blinded filtration study (4) results in a mean reduction of `r round((1-s5_means[1,5])*100)`-`r round((1-s5_means[1,6])*100)`% (see Table S5).


Ignoring data from the passive control group (3) for the meta-analysis, leads to a mean reduction in mortality odds of `r round((1-s5_means[1,7])*100)`-`r round((1-s5_means[1,8])*100)`% (see Table S5).

When those who receive spring protection in year 2 are included in the treatment group (15) for the meta-analysis, the estimated mean reduction in mortality ranges from `r round((1-s5_means[1,9])*100)`-`r round((1-s5_means[1,10])*100)`% (see Table S5).

Dropping studies where the water treatment intervention was combined with the provision of cookstoves (6) or other hygiene and sanitation interventions (7) leads to significant OR estimates, with a mean reduction in mortality odds around 30% (Frequentist OR `r s5_means[1,11]`, Bayes OR `r s5_means[1,12]`), see Table S5.

The results are plotted in Figure S1. We find that the mean reduction in mortality odds ranges from `r freq_estimates_duration %>% summarise((1 - max(mean))*100) %>% round`% to `r freq_estimates_duration %>% summarise((1 - min(mean))*100) %>% round`%. Estimates remain significant at 5% until eleven shortest studies are excluded, but the trend is mild: all estimates are significant at the 10% level.

We also recalculated the ORs using a fixed continuity correction of **0.25** and an assumption of an ICC of **1.8%** (in the clustered RCTs; see Section 2, “Obtaining input odds ratios for the model”) for the **11** studies for which we don’t have individual-level data. For the remaining **seven** studies we used the Bayesian model to appropriately account for clustering and no continuity corrections. With these inputs, a Bayesian model estimates a `r round((1-s5_means[1,13])*100)`% reduction in mortality odds (95% CrI `r s5[2,13] %>% str_remove_all("[(]") %>% str_remove_all("[)]")`), while a frequentist model estimates a `r round((1-s5_means[1,14])*100)`% reduction (95% CrI `r s5[2,14] %>% str_remove_all("[(]") %>% str_remove_all("[)]")`).


We conducted an additional check of whether short studies may be unduly impacting the model. We started from **11** studies in the dataset that include one year or more of follow-up data and fit the frequentist model. Then, we considered a hypothetical short study of 13 weeks (3 months), where the death risk is supposed to (crudely) approximate event rates in the dataset, 0.4%, and the size of the control arm is same as average size of control in the dataset, 1178. We assumed 1:1 randomization and that the true OR is the same as in the model of 10 long studies (0.80). We then simulated a growing number of short studies, 1, 2, 3, …, 10, in each case conducting 100 replications. We examined the behavior of mean and 95% intervals. Predictably, the mean was not affected and the intervals shrank only slightly: in the model of only 10 long studies the 95% interval was **60.6% to 93.0%**. In the model with 10 long and 10 simulated short studies the 95% interval was **66.9% to 95.2%** (averaged over 100 replications). This suggests that including short studies has a negligible impact on precision of the estimate, unless they have high event rates.


#### Fixed and random effects Bayesian models

```{r fe vs re}
load(here("output/stan/bayesian-ma-models.Rdata"))

bayes_or_fe <- 
  oeb(bg_loo_full$full_model) %>% 
  round(2) 
```

We fitted both fixed-effects and random-effects models. Under a fixed effect Bayesian logit model the reduction in odds was `r (1-bayes_or_fe["mean"])*100`% (OR `r bayes_or_fe["mean"]`, 95% CrI `r bayes_or_fe["lower"]`, `r bayes_or_fe["upper"]`), compared to `r bayes_reduction`% under the random effects model. Using a leave-one-study-out cross-validation (LOO CV) procedure, the expected log predictive density (ELPD) for the partial pooling model was `r bg_loo$elpd %>% round(1)` (with SE of `r bg_loo$se %>% round(1)`) and for the full pooling model `r bg_loo_full$elpd %>% round(1)` (SE of `r bg_loo_full$se %>% round(1)`). This suggests no significant differences in the out-of-sample performance of both models, with slight preference for the full pooling model.

## 4 Cost-effectiveness analysis


Table S6 presents estimates of the magnitude of benefits of a hypothetical global program to deliver water treatment through MCH. MCH programs are most relevant in settings where piped water (which can be treated using inline chlorination) is not available. As such table S6 considers the cost of a program targeting households in low- and lower-middle income countries who do not yet have access to piped water. 

2.2 billion people around the world do not have access to safely managed drinking water services (9), see row 1. This number is similar in magnitude to the global estimates from other studies (20, 21). Of this population approximately 220 million are children under the age of 5 (22, 23). 

The under-5 mortality rate among populations without access to safe drinking water implies **2.3** million deaths per year in absence of water treatment. Based on the expected effect of water treatment on child mortality in a new implementation, and assuming a 26 percentage point increase in chlorination, as observed in Malawi (18), we estimate that a program that targets this population would save around `r (plyr::round_any(s6_total$u5_saved, 50) * 1000) %>% format(scientific  = FALSE, big.mark = ",")` under-five lives at a cost of approximately USD `r plyr::round_any(s6_total$coupons_cost, 50)` million each year (Table S6). 

As noted, we include these estimates not to recommend these particular approaches to water treatment, as other approaches may be better suited to particular contexts. However, since this could be achievable virtually anywhere, it serves as a lower bound. Additionally, as the developing world becomes increasingly urban, our estimates potentially can be applied to improving access to clean water through piped water systems. 


## 5 Publication bias

```{r pub bias}
# Andrews-kasy
source(here("code/publication_bias/andrews-kasy.R"))

pub_bias_2 <- round(pub_bias, 2)
```


### Publication bias on diarrhea outcome

Using the same dataset of diarrhea outcomes as in section 1, but with diarrhea outcomes from several studies included in this meta-analysis added to the original dataset, for a total of `r nrow(df_diarrhea)` observations (see Materials and Methods section 1 just above for details). For simplicity, we assumed multiple observations from the same publication are independent. The outcome variable was risk ratio for diarrhea in children under-5, same as in (8).

We created funnel plots (Figure S4) and estimated Egger’s tests for funnel plot asymmetry for all studies and studies that include chlorination interventions only. We rejected the hypothesis of asymmetry in both cases (p-value = `r bias_clean_fit$pval %>% round(3)`, p-value = `r bias_chlorine_fit$pval %>% round(3)` respectively), which matches the result in Wolf et al., who also found no evidence of funnel plot asymmetry across water interventions (p-value = 0.8).

We also used Andrews and Kasy’s publication bias correction technique (12) on the joint dataset. We assumed symmetric publication bias cut-off around z = |1.96| and the meta-study replication method. For this result we included all data points. The distribution of intervention effects, adjusted for publication bias and assumed to be normal, has a (hyper)mean of `r pub_bias_2["diarrhea_pe", "mu"]` (SE = `r pub_bias_2["diarrhea_se", "mu"]`) and (hyper)SD of `r pub_bias_2["diarrhea_pe", "tau"]` (SE = `r pub_bias_2["diarrhea_se", "tau"]`). The relative probability of publication (between studies with |z| less/more than 1.96), beta_p, is `r pub_bias["diarrhea_pe", "beta_p"] %>% round(3)`, with a standard error of `r pub_bias["diarrhea_se", "beta_p"] %>% round(3)`. That is, insignificant results have virtually the same estimated probability of getting published as significant results. 

Repeating the same method for the subset of studies that considered chlorination, the results are noisier, with a (hyper)mean of `r pub_bias_2["chlorine_pe", "mu"]` (SE = `r pub_bias_2["chlorine_se", "mu"]`), a (hyper)SD of `r pub_bias_2["chlorine_pe", "tau"]` (SE = `r pub_bias_2["chlorine_se", "tau"]`), and beta_p of `r pub_bias["chlorine_pe", "beta_p"] %>% round(3)` (SE = `r pub_bias["chlorine_se", "beta_p"] %>% round(3)`). Here, too, we cannot reject the hypothesis that relative publication probability is equal to 1.


### Relationship between mortality and diarrhea

As mentioned, point estimates and z-values for diarrhea outcome are available for `r df_studies %>% dplyr::filter(!is.na(effect_estimate_on_diarrhea)) %>% nrow` observations, including `r df_studies %>% dplyr::filter(!is.na(effect_estimate_on_diarrhea), trial_name %in% df_all_ma_adj$trial_name) %>% nrow` out of `r n_included` studies with mortality data. We tested (using a logistic model) if availability of mortality data in a given study depends on (1) point estimate of diarrhea effect in that study, (2) absolute z-value of the diarrhea effect exceeding 1.96. In both cases the diarrhea effect was measured as log(RR). We found no evidence of either association, with p-values of `r summary(logit_1) %>% pluck(coefficients) %>% as.data.frame() %>% slice(2) %>% pull('Pr(>|z|)') %>% round(2)` and `r summary(logit_2) %>% pluck(coefficients) %>% as.data.frame() %>% slice(2) %>% pull('Pr(>|z|)') %>% round(2)` respectively.
